Predicting the effects of character displacement on food-web dynamics

### 2019-11-07 manuscript_ECD_model.tex
tidy up and clarify files in repo

\documentclass[11pt,]{article}
\usepackage[]{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            colorlinks=true,
            linkcolor=black,
            citecolor=black,
            urlcolor=black,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{amnat}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

%%% From AmNat_MS_template.tex
%\documentclass[11pt]{article}
%\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
%\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.5} % 1.7
%\usepackage[utf8]{inputenc}
\usepackage{lineno}

% controlling text size in figure captions
\usepackage{caption}
\captionsetup[figure]{font=small}

%%%%%%% Sections not currently working properly
%\usepackage{titlesec}
%\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
%\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
%\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
%\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

%%% My added packages
%\usepackage{amsmath} % for matrices
%\usepackage{graphicx} % for organizing figures
%\usepackage{lmodern} % for tilde
%\usepackage[T1]{fontenc} % for tilde

%%%%%%%%% Previous - likely useful though
%\usepackage{setspace}\doublespacing
%\usepackage{float}
%\let\origfigure\figure
%\let\endorigfigure\endfigure
%\renewenvironment{figure}[1][2] {
%    \expandafter\origfigure\expandafter[H]
%} {
%    \endorigfigure
%}

@@

\noindent 1. University of Zurich, Department of Evolutionary Biology
and Environmental Studies, 8057 Zurich, Switzerland;

\noindent 2. University of British Columbia, Department of Zoology,
Vancouver, BC V6T 1Z4, Canada;

\(^\ast\) Corresponding author; e-mail:
\href{mailto:matthew.barbour@ieu.uzh.ch}{\nolinkurl{matthew.barbour@ieu.uzh.ch}}

@@

@@

\begin{figure}
\centering
\includegraphics{ECD_model_foraging_scenarios}
\caption{\label{fig:foraging_scenarios}\textbf{Ecological foraging
scenarios.} I examined whether the effect of ecological character
displacement on food-web dynamics depended on whether consumers competed
for resources that occurred in the same vs.~different habitats. The
width of each arrow corresponds to initial attack rates
(\emph{a}\textsubscript{ij}) of consumer \emph{j} on resource \emph{i}.
Note that \emph{C}\textsubscript{1} was pre-adapted to
\emph{R}\textsubscript{1} (\emph{a}\textsubscript{11} \textgreater{}
\emph{a}\textsubscript{21}), while \emph{C}\textsubscript{2} was a
mirror image, being pre-adapted to \emph{R\textsubscript{2}}
(\emph{a}\textsubscript{22} \textgreater{} \emph{a}\textsubscript{12}).
In each scenario, I assumed consumer feeding rates increased linearly
with resource abundance. I also relax this assumption and consider a
more realistic functional response when resources occurred in different
habitats.}
\end{figure}

@@

To gain insight to the eco-evolutionary feedback generated by character
displacement, I conducted simulations using an Adaptive Dynamics
approach. Specifically, after letting consumer and resource abundances
reach a steady state, I created a mutant consumer by randomly choosing
one and modifying its attack rate on one resource by either subtracting
or adding a small constant (0.01 in the following simulations) with
equal probability. The mutant's attack rate on the other resource was
determined by a tradeoff, such that
\(\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1\),
where \(A\) is the total investment in attack rates and \(n\) describes
the shape of the tradeoff \citep{Sargent2006}. This function has the
useful property that it differentiates between cases where intermediate
combinations of \(a_{\text{1},j}\) and \(a_{\text{2},j}\) are higher
than the extremes (when \(n>1\), green line in fig. \ref{fig:tradeoff})
or, conversely, where the two extremes are higher than intermediate
investments (when \(n<1\), orange line fig. \ref{fig:tradeoff}). When
\(n=1\), the tradeoff function is linear, and all combinations of
\(a_{\text{1},j}\) and \(a_{\text{2},j}\) have the same total attack
rate (blue line in fig. \ref{fig:tradeoff}). Assuming the mutant
consumer was rare, I then determined whether the mutant had higher
relative fitness than the resident consumer, and thus could invade and
replace the resident consumer. If the mutant was able to invade, I
updated the attack rate of the resident consumer to the mutant attack
rate and allowed consumer and resource abundances to reach a steady
state. I then repeated the simulation up to 10,000 times, which was
sufficient for consumers to either reach an evolutionary stable strategy
\citep[ESS,][]{Smith1973} or an evolutionary limit (e.g.
\(\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}\) is constrained to a
maximum of 1 and minimum of 0). Unless otherwise noted, I conducted
simulations with the following parameter values: \emph{r} = 1; \emph{K}
= 4; \emph{e} = 0.8; \emph{m} = 1; \emph{A} = 2; \emph{h} = 0.4; and
\(w_{11} = w_{22} = 0.6\). I set an initial value of
\(a_{11} = a_{22} = 1.2\), while \(a_{12}\) and \(a_{21}\) depended on
the value of \emph{n}. I set initial consumer and resource abundances
to: \(R_1 = R_2 = 2\); \(C_1 = C_2 = 1\). All simulations were conducted
in R \citep{R} and the code to reproduce these simulations is publically
available on GitHub (\url{https://github.com/mabarbour/ECD_model}).

\begin{figure}
\centering
\includegraphics{attack_tradeoffs.pdf}
\caption{\label{fig:tradeoff}\textbf{Evolutionary tradeoffs in consumer
attack rates.} In each foraging scenario, I explored the effects of
three different tradeoffs: intermediate combinations of attack rates
(\emph{a}\textsubscript{1,\emph{j}}, \emph{a}\textsubscript{2,\emph{j}})
are higher than the extremes (green line, \emph{n} \textgreater{} 1);
extreme combinations of attack rates are higher than intermediate
investments (orange line, \emph{n} \textless{} 1); and all combinations
of attack rates have the same total attack rate (blue line, \emph{n} =
1). Points corresponding to attack rates at the beginning of the
simulation for \emph{C}\textsubscript{1}, which was pre-adapted to
\emph{R}\textsubscript{1} (\emph{a}\textsubscript{11} \textgreater{}
\emph{a}\textsubscript{12}).}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Figure_3.pdf}
\caption{\label{fig:plot_fig3}\textbf{Effect of character displacement
on food-web dynamics under different evolutionary tradeoffs and foraging
scenarios.} Lines show predicted values when both consumers and
resources are present. Different line colors correspond to different
tradeoffs in attack rates (green, \emph{n} = 1.15; blue, \emph{n} = 1;
orange, \emph{n} = 0.85). Large circles (two consumers) and triangles
(one consumer) correspond to the end points of the eco-evolutionary
simulation for \emph{C}\textsubscript{1}, whereas as small shapes
correspond to the starting points (only in stability panels). In both
foraging scenarios, feeding rates increase linearly with resource
abundance, but the equation for effective attack rate is different.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{bifurcation_plot.pdf}
\caption{\label{fig:bifur_plot}\textbf{Character divergence creates an
unstable food web}. Lines illustrate the effects of character divergence
across the range of specialization, while the points are the results of
an eco-evolutionary simulation. Note that I increased the total
investment in attack rates (\emph{A} = 3.3) to create a scenario that
could result in an unstable food web.}
\end{figure}

@@

This work was inspired by discussions with Seth Rudman, Dolph Schluter,
Ben Gilbert, and Kevin McCann. Sally Otto provided much help in early
analyses of the mathematical model. I would not have been able to take
the theory as far as I did without her guidance and encouragement. For
funding support, I thank the University of British Columbia (Four-Year
Fellowship to M.A.~Barbour), NSERC (Discovery grant to Greg Crutsinger),
and the Swiss National Science Foundation (grant 31003A\_160671 to Jordi
Bascompte).

\bibliography{references}

>>>>

\documentclass[11pt,]{article}
\usepackage[]{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            colorlinks=true,
            linkcolor=black,
            citecolor=black,
            urlcolor=black,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{EcolLett}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{fullpage}
\linespread{1.5}
\usepackage{lineno}
\usepackage{caption}
\captionsetup[figure]{font=small}

@@

\noindent 1. University of Zurich, Department of Evolutionary Biology
and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich,
Switzerland;

\noindent 2. University of British Columbia, Department of Zoology, 2212
Main Mall, Vancouver, BC V6T 1Z4, Canada;

\(^\ast\) Corresponding author; e-mail:
\href{mailto:matthew.barbour@ieu.uzh.ch}{\nolinkurl{matthew.barbour@ieu.uzh.ch}},
phone: +41 44 635 61 63

\emph{Statement of authorship}: MAB conceived the study, derived
analytical solutions, performed simulations, and wrote the paper.

\emph{Data accessibility}: No new data were used. All code used to
reproduce this manuscript and simulations within are publically
available on GitHub (\url{https://github.com/mabarbour/ECD_model}) and
have been archived with Zenodo (INSERT DOI).

@@

\emph{Manuscript type}: Letter.

\emph{Number of words in abstract}: 143; \emph{Number of words in main
text}: 4086; \emph{Number of words in text boxes}: NA; \emph{Number of
references}: 45; \emph{Number of figures}: 4; \emph{Number of tables}:
0; \emph{Number of text boxes}: 0

@@

\begin{figure}
\centering
\includegraphics{Fig_1_ForagingScenarios}
\caption{\label{fig:foraging_scenarios}\textbf{Ecological foraging
scenarios.} I examined whether the effect of ecological character
displacement on food-web dynamics depended on whether consumers competed
for resources that occurred in the same vs.~different habitats. The
width of each arrow corresponds to initial attack rates
(\emph{a}\textsubscript{ij}) of consumer \emph{j} on resource \emph{i}.
Note that \emph{C}\textsubscript{1} was pre-adapted to
\emph{R}\textsubscript{1} (\emph{a}\textsubscript{11} \textgreater{}
\emph{a}\textsubscript{21}), while \emph{C}\textsubscript{2} was a
mirror image, being pre-adapted to \emph{R\textsubscript{2}}
(\emph{a}\textsubscript{22} \textgreater{} \emph{a}\textsubscript{12}).
In each scenario, I assumed consumer feeding rates increased linearly
with resource abundance. I also relax this assumption and consider a
more realistic functional response when resources occurred in different
habitats.}
\end{figure}

@@

To gain insight to the eco-evolutionary feedback generated by character
displacement, I conducted simulations using an Adaptive Dynamics
approach. Specifically, after letting consumer and resource abundances
reach a steady state, I created a mutant consumer by randomly choosing
one and modifying its attack rate on one resource by either subtracting
or adding a small constant (0.01 in the following simulations) with
equal probability. The mutant's attack rate on the other resource was
determined by a tradeoff, such that
\(\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1\),
where \(A\) is the total investment in attack rates and \(n\) describes
the shape of the tradeoff \citep{Sargent2006}. This function has the
useful property that it differentiates between cases where intermediate
combinations of \(a_{\text{1},j}\) and \(a_{\text{2},j}\) are higher
than the extremes (when \(n>1\), green line in fig. \ref{fig:tradeoff})
or, conversely, where the two extremes are higher than intermediate
investments (when \(n<1\), orange line fig. \ref{fig:tradeoff}). When
\(n=1\), the tradeoff function is linear, and all combinations of
\(a_{\text{1},j}\) and \(a_{\text{2},j}\) have the same total attack
rate (blue line in fig. \ref{fig:tradeoff}). Assuming the mutant
consumer was rare, I then determined whether the mutant had higher
relative fitness than the resident consumer, and thus could invade and
replace the resident consumer. If the mutant was able to invade, I
updated the attack rate of the resident consumer to the mutant attack
rate and allowed consumer and resource abundances to reach a steady
state. I then repeated the simulation up to 10,000 times, which was
sufficient for consumers to either reach an evolutionary stable strategy
\citep[ESS,][]{Smith1973} or an evolutionary limit (e.g.
\(\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}\) is constrained to a
maximum of 1 and minimum of 0). Unless otherwise noted, I conducted
simulations with the following parameter values: \emph{r} = 1; \emph{K}
= 4; \emph{e} = 0.8; \emph{m} = 1; \emph{A} = 2; \emph{h} = 0.4; and
\(w_{11} = w_{22} = 0.6\). I set an initial value of
\(a_{11} = a_{22} = 1.2\), while \(a_{12}\) and \(a_{21}\) depended on
the value of \emph{n}. I set initial consumer and resource abundances
to: \(R_1 = R_2 = 2\); \(C_1 = C_2 = 1\). All simulations were conducted
in R \citep{R} and the code to reproduce these simulations is publically
available on GitHub (\url{https://github.com/mabarbour/ECD_model}) and
have been archived with Zenodo (INSERT DOI).

\begin{figure}
\centering
\includegraphics{Fig_2_Tradeoffs.pdf}
\caption{\label{fig:tradeoff}\textbf{Evolutionary tradeoffs in consumer
attack rates.} In each foraging scenario, I explored the effects of
three different tradeoffs: intermediate combinations of attack rates
(\emph{a}\textsubscript{1,\emph{j}}, \emph{a}\textsubscript{2,\emph{j}})
are higher than the extremes (green line, \emph{n} \textgreater{} 1);
extreme combinations of attack rates are higher than intermediate
investments (orange line, \emph{n} \textless{} 1); and all combinations
of attack rates have the same total attack rate (blue line, \emph{n} =
1). Points corresponding to attack rates at the beginning of the
simulation for \emph{C}\textsubscript{1}, which was pre-adapted to
\emph{R}\textsubscript{1} (\emph{a}\textsubscript{11} \textgreater{}
\emph{a}\textsubscript{12}).}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Fig_3_MacArthur_LawlorSmith.pdf}
\caption{\label{fig:plot_fig3}\textbf{Effect of character displacement
on food-web dynamics under different evolutionary tradeoffs and foraging
scenarios.} Lines show predicted values when both consumers and
resources are present. Different line colors correspond to different
tradeoffs in attack rates (green, \emph{n} = 1.15; blue, \emph{n} = 1;
orange, \emph{n} = 0.85). Large circles (two consumers) and triangles
(one consumer) correspond to the end points of the eco-evolutionary
simulation for \emph{C}\textsubscript{1}, whereas as small shapes
correspond to the starting points (only in stability panels). In both
foraging scenarios, feeding rates increase linearly with resource
abundance, but the equation for effective attack rate is different.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Fig_4_Bifurcation.pdf}
\caption{\label{fig:bifur_plot}\textbf{Character divergence creates an
unstable food web}. Lines illustrate the effects of character divergence
across the range of specialization, while the points are the results of
an eco-evolutionary simulation. Note that I increased the total
investment in attack rates (\emph{A} = 3.3) to create a scenario that
could result in an unstable food web.}
\end{figure}

@@

This work was inspired by discussions with Seth Rudman, Dolph Schluter,
Ben Gilbert, and Kevin McCann. Sally Otto provided much help in early
analyses of the mathematical model. I would not have been able to take
the theory as far as I did without her guidance and encouragement. Matt
Osmond, Jean Gibert, and Seth Rudman provided critical feedback on an
earlier draft. For funding support, I thank the University of British
Columbia (Four-Year Fellowship to M.A.~Barbour), NSERC (Discovery grant
to Greg Crutsinger), and the Swiss National Science Foundation (grant
31003A\_160671 to Jordi Bascompte).

\bibliography{ReferencesAbbrev}
